# Getting started

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

Load the **ggplot2** library and read in the example dataset we'll be using for most of these plots. This is a timeseries of detections of different whale species collected by an ocean glider off southern Nova Scotia, Canada, in the fall of 2017. 

```{r, results='hide'}

# the ggplot library
library(ggplot2)

# the dplyr library (for occasional data manipulation)
library(dplyr)

# daily counts of whale daily detections
df = readRDS(url('https://hansenjohnson.org/files/data/detections_daily.rds'))
```

## Plotting

Let's make a bar plot of how the number of detections per day (`n`) changed over time (`date`). We'll specify that we're going to use `df` as our input data, that the variable `date` should be mapped to the x axis and the variable `n` should be mapped to the y axis, and that we'd like to use a column (`col`) geometry. Here's how to do that: 

```{r}
ggplot(data = df, mapping = aes(x=date,y=n))+
  geom_col()
```

There are a few things to note here:

- every ggplot2 plot must start with the `ggplot()` function  
- the `+` is used to connect multiple functions associated with the given plot  
- the `data = ` argument is used to refer to the input table  
- the `mapping = ` argument is used to refer to the mapping aesthetics, which are specified using the `aes()` function  
- the geometry is specified using the `geom_col()` function.    

You will also see this written as follows:
```{r}
ggplot()+
  geom_col(data = df, mapping = aes(x=date,y=n))
```

This produces identical results. I prefer using this method because it makes things easier down the road when things get more complicated and you're using multiple geometries to map different aesthetics from different datasets.

Now let's get a little more sophisticated and add use some color to distinguish between different species (`species`). That is as easy as setting the `fill=species` within the aesthetics mapping (`aes()`).

```{r}
ggplot()+
  geom_col(data = df, mapping = aes(x=date,y=n,fill=species))
```

>**Note** - the default ggplot colors are **NOT** colorblind friendly!

Aesthetics that are specified outside of `aes()` are applied to that geometry, but not mapped to the data. Here `color = 'black'` will make the border of all bars black, regardless of the mapping between aesthetics and data.

```{r}
ggplot()+
  geom_col(data = df, mapping = aes(x=date,y=n,fill=species), color = 'black')
```

The output of a `ggplot()` function can be stored as a variable, which can be further altered, plotted, or saved later on.

Let's build the plot again, but store it as `p` instead of plotting it
```{r}
p <- ggplot()+
  geom_col(data = df, mapping = aes(x=date,y=n,fill=species), color = 'black')
```

And now we can plot it by simply printing the variable p
```{r}
p
```

## Formatting

The default formatting options have been carefully chosen so that plots produced by ggplot2 tend to look pretty good right out of the box. Let's go through some simple, common formatting tweaks.

### Labels

Updating labels is easy. There are a number of ways to do it. I prefer to use the `labs()` function and specify the new label you would like to be associated with each aesthetic. I'll use `NULL` when I don't want a label to appear

```{r}
p + labs(x = NULL, 
         y = 'Detections per day', 
         fill = 'Species')
```

Note how I split each label onto a new line. This is optional but makes the code more readible.

### Scales

It's common to want to change the scales in some way. This is achieved using a suite of functions with the following naming convention `scale_[aesthetic]_*()`. I often do this to update the axis limits or change the default color scheme (I really don't like the default ggplot2 colors!)

Here's how to change the y axis limits and break points. I also often use the `expand = c(0,0)` arugment to remove the margins at the top and bottom of the plot area.

```{r}
p + scale_y_continuous(limits = c(0,50), breaks = seq(0,50,5), expand = c(0,0))
```

Here's how to change the color map. I recommend setting the values by hand for discrete variables, and using a viridis colormap for continuous variables. Here's a convenient way to map color values by hand using a named list.

```{r}
p + scale_fill_manual(values = c('fin' = 'grey', 
                                 'humpback' = 'blue', 
                                 'right' = 'red', 
                                 'sei' = 'white'))
```

And finally, here's how to change the time resolution on the x axis.

```{r}
p + scale_x_date(date_breaks = '3 day', date_labels = '%b %d')
```

### Theme

The overall look and feel of a plot are specified by the `theme()` function. There are also numerous default themes that can be called as follows:

- `theme_bw()`  
- `theme_classic()`  
- `theme_minimal()`  
- `theme_void()`  
- `theme_dark()`  

Here's how our original plot looks with the `theme_bw()` theme applied

```{r}
p+theme_classic()
```

You can also use arguments within the generic `theme()` function to change *literally* everything about how a plot looks. Here's a quick example of how to turn the gridlines off:

```{r}
p+theme(panel.grid = element_blank())
```

### Putting it all together

Let's build a nicely formatted plot from scratch

```{r}
# start ggplot
p_bar = ggplot()+
  
  # specify geometry, data, and aesthetic mapping
  geom_col(data = df, mapping = aes(x=date,y=n,fill=species), color = 'black')+
  
  # define fill colors to use for each species
  scale_fill_manual(values = c('fin' = 'grey', 
                               'humpback' = 'blue', 
                               'right' = 'red', 
                               'sei' = 'white'))+
  
  # update the y axis limits and break points
  scale_y_continuous(limits = c(0,50), breaks = seq(0,50,5), expand = c(0,0))+
  
  # update x axis for more regular date breaks
  scale_x_date(date_breaks = '3 day', date_labels = '%b %d')+
  
  # write nice labels
  labs(x = NULL, y = 'Detections per day', fill = 'Species')+
  
  theme_classic()

# print plot  
p_bar
```

### Saving

Saving a plot in **ggplot2** is done using the `ggsave()` function. Here's how I'd do that with the example plot:

```{r, eval=FALSE}
ggsave(filename = 'bar.png', plot = p_bar, 
       width = 5, height = 5, units = 'in', dpi = 300)
```
